Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-17 linear_adia.tex
write up edits

\subsection{Adiabatic perturbations} 
We now extend the above discussion to vertically isothermal disks
($\Gamma=1$) with $\gamma\neq1$ and $t_c\to\infty$, i.e. adiabatic
evolution. We eliminate variables in favor of $\delta v_z$ and
$\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
&\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
  v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
  - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
& -\sigma^2\delta v_z = \gamma c_s^2 \frac{d\Delta}{dz} +
c_s^2\frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
\left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
{\bf note: incompressible mode, set delta to zero. only works for
  constant gravity.}

We again make the low-frequency approximation
and ignore the vertical dependence of $\Omega$ and $\kappa$ except
where their vertical derivatives appear explicitly in the above
equations. We thus make the replacement $D\to \Omega_k^2$, as before
for isothermal perturbations. Here, we also use
the fact that in the global disk,
\begin{align}
  r\frac{\p \Omega^2}{\p z} = - \frac{\p\ln\rho}{\p z}\frac{\p
    c_s^2}{\p r} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
\end{align}
where the second equality follows from the power-law prescription of
the sound-speed for $\Gamma=1$.

@@

Note that the term with $\ln\rho^{\prime 2}$ in the integrand (third
term on the RHS) represents stabilization by the vertical entropy
gradient when there is no vertical shear. It gives an imaginary
contribution to $\hat{\sigma}$ when $q\neq0$, but this is a factor
$|\epsilon/\hat{k}|$ smaller than the real (stabilizing) contribution,
since $\epsilon\ll 1$ for a thin disk and we (usually) consider
$\hat{k}\gg1$. Thus, instability is expected to be due to the 
last term on the RHS of Eq. \ref{adia_integral}.

\subsubsection{Explicit solutions in the thin-disk limit}
To simplify further, we consider the problem in the thin-disk limit so
that $\ln \rho \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes 
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z,
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. We require
the vertical kinetic energy density $\rho|\delta v_z|^2$ to remain
finite.  Then assuming
$g(\zhat)$ is a polynomial, we require 
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}

Inserting the ansatz Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin}
gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
      \alpha\right)g
    +\underbrace{\left(\alpha^2 - \alpha A - C\right)}_\text{set to zero}\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$ vanish. That
is,
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).
\end{align} 
We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
%Recalling that $B$ depends $\hat{\sigma}$, Eq. \ref{adia_disp} gives
%the eigenfrequency as a function of disk paramters.

\subsubsection{The case of $\gamma=2$}
In principle, we can solve Eq. \ref{adia_disp} for the mode
frequencies and growth rates. This is complicated in 
general, but simplifies for $\gamma=2$, for which the quantity
$A^2+4C$ is real. In this case we find
\begin{align}
  \hat{\nu}^2 =
  \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
    \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
  &\phantom{=}
  \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
\end{align}
In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
have
\begin{align}
  \hat{\nu}^2 = \frac{\epsilon^2
    q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
\end{align}
and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
\begin{align}
  \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
    q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
  % & \leq \frac{|\epsilon
  % q|}{2\times3^{3/4}}\notag.
\end{align}
Eq. \ref{gam2_max_growth} show that growth rates decrease with $M$,
and there exists a maximum possible growth rate at $M=0$. This is
despite the fact that in the thin-disk approximation, for which
$d\Omega^2/dz\propto z$, there is infinite vertical shear
available.

\subsubsection{Comparison with $\gamma=1$}
We can compare growth rates obtained for $\gamma=2$ to those for
$\gamma=1$ in the previous section. If we compare growth rates at their 
respective optimum radial wavelengths, we find
that for $\gamma=2$ the \emph{maximum} possible growth rate, occuring
at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
rate for $\gamma=1$. That is,
\begin{align}
  &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
      q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
  &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
           & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
\end{align}

>>>>

\subsection{Adiabatic perturbations} 
We now extend the above discussion to general $\gamma$, but in the
adiabatic limit ($t_c\to\infty$). We eliminate variables from 
Eq. \ref{lin_mass}---\ref{lin_energy} in favor of $\delta v_z$ 
and $\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
  &\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
    v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
    - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
  & -\sigma^2\delta v_z = \gamma c_s^2 \frac{d\Delta}{dz} +
  c_s^2\frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
  \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
{\bf note: incompressible mode, set delta to zero. only works for
  constant gravity.}

We again make the low-frequency approximation
and ignore the vertical dependence of $\Omega$ and $\kappa$ except
where their vertical derivatives appear explicitly in the above
equations. We thus make the replacement $D\to \Omega_k^2$, as before
for isothermal perturbations. Here, we also use the fact that in the
global disk, 
\begin{align}
  r\frac{\p \Omega^2}{\p z} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
\end{align}
for the power-law prescription of the midplane sound-speed for
$\Gamma=1$.

@@

Note that the term with $\ln\rho^{\prime 2}$ in the integrand (third
term on the RHS) represents stabilization by the vertical entropy
gradient when there is no vertical shear. It gives an imaginary
contribution to $\hat{\sigma}$ when $q\neq0$, but this is a factor
$|\epsilon/\hat{k}|$ smaller than the real (stabilizing) contribution,
since $\epsilon\ll 1$ for a thin disk and we (usually) consider
$\hat{k}\gg1$. Thus, instability is expected to associated with the 
last term on the RHS of Eq. \ref{adia_integral}.

\subsubsection{Explicit solutions in the thin-disk limit}
We can construct  solutions in the thin-disk limit, in which 
$\ln(\rho/\rho_0) \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes   
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z,
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish. Furthermore, we require 
the vertical kinetic energy density $\rho|\delta v_z|^2$ to remain
finite. Then assuming $g(\zhat)$ is a polynomial, we require 
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}
We therefore choose 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right). 
\end{align}

We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
%Recalling that $B$ depends $\hat{\sigma}$, Eq. \ref{adia_disp} gives
%the eigenfrequency as a function of disk paramters.

\subsubsection{The case of $\gamma=2$}
In principle, we can solve Eq. \ref{adia_disp} for the mode
frequencies and growth rates. This is complicated in 
general, but simplifies for $\gamma=2$, for which the quantity
$A^2+4C$ is real. In this case we find
\begin{align}
  \hat{\nu}^2 =
  \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
    \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
  &\phantom{=}
  \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
\end{align}
In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
have 
\begin{align}
  \hat{\nu}^2 = \frac{\epsilon^2
    q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
\end{align}
and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
\begin{align}
  \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
    q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
  % & \leq \frac{|\epsilon
  % q|}{2\times3^{3/4}}\notag.
\end{align}
Unlike for isothermal perturbations, Eq. \ref{gam2_max_growth} show
that growth rates decrease with $M$, and there exists a maximum
possible growth rate at $M=0$. This is despite the fact that in the
thin-disk approximation, for which $d\Omega^2/dz\propto z$, there is
infinite vertical shear available.

The acceptable solution for $\alpha$ is
\begin{align}
  \alpha = \frac{1}{2}\left[\left(1+\ii\epsilon q \hat{k}\right) -
    \sqrt{1 + \hat{k}^2\left(2-\epsilon^2 q^2\right)}\right]. 
\end{align}
Since $\epsilon\ll 1$, $\real\alpha<0$, so the vertical velocity
eventually decays for large heights. This is unlike for isothermal
perturbations, in which perturbation magnitudes increase with $|z|$.

\subsubsection{Comparison with $\gamma=1$}
We can compare growth rates obtained for $\gamma=2$ to those for
$\gamma=1$ in the previous section. If we compare growth rates at their 
respective optimum radial wavelengths, we find
that for $\gamma=2$ the \emph{maximum} growth rate, occuring
at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
rate for $\gamma=1$. That is,
\begin{align}
  &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
      q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
  &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
           & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
\end{align}
